---
output: hugodown::md_document
title: "What they forgot to tell you about SQL"
subtitle: ""
summary: "SELECT top 10 tips FROM sql_syntax"
authors: []
tags: []
categories: [data]
date: 2020-09-19
lastmod: 2020-09-19
featured: false
draft: true
aliases:

image:
  caption: "Photo by [Max Duzij](https://unsplash.com/@max_duz) on Unsplash"
  focal_point: ""
  preview_only: true

projects: [""]
---

I've always considered SQL to be "bad manager" syntax:

```
SELECT -- "go get the stuff I need"
FROM -- "oh didn't I tell you, it's from this place
UNION -- "and this other place"
LEFT JOIN -- "but you'll also need to look here"
WHERE -- "but actually throw away some of it"
GROUP BY -- "oh and please add it all up before you send"
```

This same "Oh, just one more thing..." mentality makes SQL frustrating for new data analysts to both read and write.

```{r echo = FALSE}
# Source: https://gist.github.com/gadenbuie/c83e078bf8c81b035e32c3fc0cf04ee8
render_toc <- function(
  filename, 
  toc_header_name = "Table of Contents",
  base_level = NULL,
  toc_depth = 3
) {
  x <- readLines(filename, warn = FALSE)
  x <- paste(x, collapse = "\n")
  x <- paste0("\n", x, "\n")
  for (i in 5:3) {
    regex_code_fence <- paste0("\n[`]{", i, "}.+?[`]{", i, "}\n")
    x <- gsub(regex_code_fence, "", x)
  }
  x <- strsplit(x, "\n")[[1]]
  x <- x[grepl("^#+", x)]
  if (!is.null(toc_header_name)) 
    x <- x[!grepl(paste0("^#+ ", toc_header_name), x)]
  if (is.null(base_level))
    base_level <- min(sapply(gsub("(#+).+", "\\1", x), nchar))
  start_at_base_level <- FALSE
  x <- sapply(x, function(h) {
    level <- nchar(gsub("(#+).+", "\\1", h)) - base_level
    if (level < 0) {
      stop("Cannot have negative header levels. Problematic header \"", h, '" ',
           "was considered level ", level, ". Please adjust `base_level`.")
    }
    if (level > toc_depth - 1) return("")
    if (!start_at_base_level && level == 0) start_at_base_level <<- TRUE
    if (!start_at_base_level) return("")
    if (grepl("\\{#.+\\}(\\s+)?$", h)) {
      # has special header slug
      header_text <- gsub("#+ (.+)\\s+?\\{.+$", "\\1", h)
      header_slug <- gsub(".+\\{\\s?#([-_.a-zA-Z]+).+", "\\1", h)
    } else {
      header_text <- gsub("#+\\s+?", "", h)
      header_text <- gsub("\\s+?\\{.+\\}\\s*$", "", header_text) # strip { .tabset ... }
      header_text <- gsub("^[^[:alpha:]]*\\s*", "", header_text) # remove up to first alpha char
      header_slug <- paste(strsplit(header_text, " ")[[1]], collapse="-")
      header_slug <- tolower(header_slug)
    }
    paste0(strrep(" ", level * 4), "- [", header_text, "](#", header_slug, ")")
  })
  x <- x[x != ""]
  knitr::asis_output(paste(x, collapse = "\n"))
}
```

```{r}
library(knitr)
library(DBI)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(connection = "con")
con <- dbConnect(RSQLite::SQLite(), ":memory:")

con3 <- dbConnect(duckdb::duckdb(), ":memory:", read_only = FALSE)
copy_to(con3, nycflights13::flights, "flights")
copy_to(con3, nycflights13::airports, "airports")

# 
# con2 <- dbConnect(
#   RMariaDB::MariaDB(),#RPostgres::Postgres(),
#   user = Sys.getenv("FLIGHTS_DB_USER"),
#   password = Sys.getenv("FLIGHTS_DB_PASS"),
#   dbname = "nycflights13",
#   host = "db-edu.pacha.dev"
# )
```

Thanks to Mauricio Vargas for the [test database](https://db-edu.pacha.dev/) used in this post.

## Quick Start

```{r echo = FALSE}
render_toc(
  filename = "index.Rmd",
  toc_header_name = 'Quick Start'
  )
```

Thanks to Garrick Aden-Buie for the [table of contents code](https://gist.github.com/gadenbuie/c83e078bf8c81b035e32c3fc0cf04ee8)

## Rapid Testing

### Get Schema

```{sql, connection = con3}
describe flights;
```

### Create a sample table

```{sql connection = con3}
CREATE TEMPORARY TABLE IF NOT EXISTS Order_Detail (
    invoice_id INTEGER NOT NULL,
    invoice_line INTEGER NOT NULL,
    store_id INTEGER NOT NULL,
    time_stamp DATE NOT NULL,
    product VARCHAR(8) NOT NULL,
    units INTEGER NOT NULL,
    sales NUMERIC(7 , 2 ) NOT NULL,
    cogs NUMERIC(5 , 2 ) NOT NULL
);

```

```{sql connection = con3}
INSERT INTO Order_Detail(invoice_id,invoice_line,store_id,time_stamp,product,units,sales,cogs) VALUES (1000,312,3,'2018/12/23','30',1,199.99,28.00);
```

```{sql connection = con3}
select * from Order_Detail limit 1;
```

### No `FROM` clause

```{sql connection = con}
select 10*5 as product;
```

## Defensive Programming

### Coalesce

```{sql connection = con3}
WITH const AS (SELECT 10 AS more)
SELECT sales + NULL
FROM flights
```

```{sql connection = con3}
WITH const AS (SELECT 10 AS more)
SELECT sales + coalesce(NULL, 0)
FROM order_detail as o, const
```

### Case-When Defaults

```{sql connection = con3}
WITH const AS (SELECT 10 AS more)
SELECT sales + coalesce(NULL, 0)
FROM order_detail as o, const
```

```{sql connection = con3}
SELECT case when dep_delay > 10 then '>10' when dep_delay > 0 then '>0' else '<0' end as delay, count(*)
FROM flights
group by 1
```

```{sql connection = con3}
SELECT 
  case
    when dep_delay > 10 then '>10' 
    when dep_delay > 0 then '>0' 
    when dep_delay <= 0 then '<=0'
    else 'ERROR' 
  end as delay, 
  count(*)
FROM flights
group by 1
```

## Programming

### Local Variables

With SQLite

```{sql connection = con}
WITH const AS (SELECT 10 AS more)
SELECT sales * const.more
FROM order_detail as o, const
```

Or with MySQL

```{sql connection = con2}
set @month = 1
```

```{sql connection = con2}
select count(*) tot, sum(if(month = @month, 1, 0)) jan
from flights
```

### User-Defined Functions

```{sql connection = con2}
CREATE OR REPLACE FUNCTION iif_sql(boolean, anyelement, anyelement) returns anyelement as
$body$ select case $1 when true then $2 else $3 end $body$
LANGUAGE sql IMMUTABLE;
```


```{sql connection = con2, output.var = "x"}
select year, iif_sql(year > 2011, 1, 0)
from flights
limit 10
```

```{r echo = FALSE}
x
```

```{sql connection = con2}
CREATE OR REPLACE FUNCTION add_10 (N INT)
    RETURNS INT DETERMINISTIC
    RETURN (N+1);
```



## Underutilized Functions

### Greatest / Least

### If-Else versus Case-When

```{sql, connection = con, eval = FALSE}
select iif(1 < 2, 1, 2)
```

### Faster Binning

compare to `case when`

```{sql, connection = con}
select floor(sales / 50) * 50 as bin_sales
from order_detail

```

## More Advanced Calculations

### Window Functions

### Aggregate Filters

`having`

```{sql connection = con2}
select month, count(*) as n
from flights
group by 1
having n > 28000
```

`qualify`

```{sql connection = con2}
select month, count(*) as n
from flights
group by 1
qualify mean()
```
